Multi-level Selective Deduplication for VM Snapshots in Cloud Storage

### 2012-03-01 vm_snapshot_sim/2012background.tex

Data deduplication techniques can eliminate redundancy globally among different files from different users. 
Backup systems have been developed to use content hash (finger prints) to identify duplicate 
content~\cite{venti02,Rhea2008}.
%,NGmiddleware2011}. 
Today's commercial data backup systems (e.g. from EMC and NetApp)
%\cite{emc_avamar}\cite{datadomain_whitepaper} 
use a variable-size chunking algorithm to detect duplicates in file data~\cite{similar94,hydrastor09}.
%Chunking divides a data stream into variable length chunks, it has been used to conserve bandwidth\cite{lbfs01}, 
%search and index documents in large repositories\cite{bhag07}, scalable storage systems\cite{hydrastor09}, 
%store and transfer large directory trees efficiently and with high reliability\cite{jumbo07}.
To chunk a file, starting from the first byte, its contents as seen through a fixed-sized (overlapping) 
sliding window are examined. At every position of the window, a fingerprint of its 
contents, $f$, is computed using hashing techniques such as Rabin fingerprints~\cite{rabin81}. 
When this fingerprint meets a certain criteria,
% such as $f mod D = r$ where $D$, the divisor, 
%and $r$ are predefined values; 
the position of the corresponding window defines the boundary of the chunk. 
This process is repeated until the entire file has been completely broken down into chunks. Next, 
a cryptographic hash or chunk ID of the chunk is computed using a technique such as MD5 or SHA.
After a file is chunked, the index containing the chunk IDs of backed up chunks 
is queried to determine duplicate chunks. New chunks are written to disk and the 
index is updated with their chunk IDs. A file recipe containing all the information 
required during reconstruction is generated. The index also contains some metadata 
about each chunk, such as its size and disk location.
%The compression rate obtained by deduplication depends on the inherent content overlapping degree in a dataset, 
%the average size of chunks, and the chunking method\cite{poli04}. 
%In general, smaller chunks yield better deduplication.

Data deduplication has been applied as  a post-processing operation for offline removal
of redundant content in the file system data management once backup data is written in a temporary stage area.
Dynamic inline deduplication is an effort to improve storage efficiency  
by detecting and removing  duplicates before data is written to the backup storage,
thus reduces extra disk space required to hold temporary data. 
This approach requires a fingerprint  comparison of the new content chunks with
previously stored fingerprints. 
As data grows to be big, content chunk lookup for fingerprint comparison from disk storage
becomes too slow to be scalable.
%and searching such disk
%However, unless some form of locality or similarity is exploited, inline, chunk-based deduplication, 
%when done at a large scale faces what has been termed the disk bottleneck problem: to facilitate fast chunk ID lookup, 
%a single index containing the chunk IDs of all the backed up chunks must be maintained. 
%As the backed up data grows, the index overflows the amount of RAM available and must be paged to disk. 
%Without locality, the index cannot be cached effectively, and it is common for nearly 
%every index access to require a random disk access. This disk bottleneck severely limits deduplication throughput.
Several techniques have been proposed to speedup searching of duplicate 
content~\cite{bottleneck08,extreme_binning09,sparseindex09}. For example,  
Zhu et al.~\cite{bottleneck08} tackle it 
by using an in-memory Bloom filter and prefetch groups of chunk IDs that are likely to be 
accessed together with high probability. It takes significant memory resource for filtering.
NG et al.~\cite{ NGmiddleware2011}  use  
a related filtering technique for integrating deduplication in Linux  file system and the memory
consumed is up to 2GB for a single machine. That is still too big in our context discussed below. 
%Lillibridge et al.~\cite{sparseindex09} break list of chunks 
%into large segments, the chunk IDs in each incoming segment are sampled and the segment is 
%deduplicated by comparing with the chunk IDs of only a few carefully selected backed up segments. 
%These are segments that share many chunk IDs with the incoming segment with high probability.
%Deepavali et al.~\cite{extreme_binning09}  use signature-based file similarity  and group similar files
%into the same physical location (bins) to deduplicate against each other.

@@

There are multiple choices in designing a backup architecture  for VM snapshots.
We discuss the following design options with a consideration on their strengths and weakness.
\begin{enumerate}
\item  {\em An external and dedicated backup storage system.} 
In this architecture setting, a separate backup storage system using
the standard backup and deduplication techniques can be deployed~\cite{bottleneck08,extreme_binning09,sparseindex09}. 
This system is attached to the cloud network and every machine can periodically transfer snapshot data to 
the attached backup system. 
A key weakness of this approach is communication bottleneck between a large number of machines
in a cloud to this centralized  service.
Another weakness is that the cost of allocating separate resource for dedicated backup  can be expensive.
Since most of backup data is not used eventually, CPU and memory resource in such a backup cluster may not be fully utilized.
\item {\em A decentralized and co-hosted backup system with full deduplication.}
In this option, the backup system runs on an existing set of cluster machines. 
The advantage is that there is no separate and dedicated cost for backup.
% and a distributed storage architecture for backup allows a possible exploitation of  data locality between
%the source of data and storage location of backup data. 
The disadvantage is that 
%the backup service would compete CPU, memory, and disk resources with the other cloud services.
even such a  backup service may only use  a fraction of the existing disk storage, 
fingerprint-based search does require a significant amount of memory for fast lookup of identical content chunks.
This competes memory resource  with the existing applications.
%Decentralized deduplication is studied in ~\cite{Clements2009} and the focus is on
%block-level copy-on-write and compare-by-value techniques.

@@

With these considerations in mind, we propose a decentralized backup architecture with multi-level and selective 
deduplication. This 
backup service is hosted   in the existing set of cloud machines and resource usage is controlled
with a minimal impact to the existing applications.
The duplication process is first conducted among data within each VM
and then is conducted  across VMs when needed to 
utilize a common dataset.  
Given the concern that searching duplicates across  VMs  is a global feature which can affect parallel performance
and complicate failure management,
we would like to use a small common dataset while still maintaining a cost-effective deduplication ratio.
For this purpose, we exploit the data characteristics of snapshots and only store most popular data.
%
%in the  our studies show that there are a large amount of data unmodified in VM snapshots
%and shared among many users (e.g. OS data). The sharing pattern follows a zip-like distribution.
%With this characteristic, we can store a small amount of popular data items which coverage a large amount of
%snapshot data blocks. 
%We discuss our design and data analysis in the next section.

>>>>

Data deduplication techniques can eliminate redundancy globally among different files from different users. 
Backup systems have been developed to use content hash (finger prints) to identify duplicate 
content~\cite{venti02,Rhea2008}.
%,NGmiddleware2011}. 
Today's commercial data backup systems (e.g. from EMC and NetApp)
%\cite{emc_avamar}\cite{datadomain_whitepaper} 
use a variable-size chunking algorithm to detect duplicates in file data~\cite{similar94,hydrastor09}.
%Chunking divides a data stream into variable length chunks, it has been used to conserve bandwidth\cite{lbfs01}, 
%search and index documents in large repositories\cite{bhag07}, scalable storage systems\cite{hydrastor09}, 
%store and transfer large directory trees efficiently and with high reliability\cite{jumbo07}.
\comments{
To chunk a file, starting from the first byte, its contents as seen through a fixed-sized (overlapping) 
sliding window are examined. At every position of the window, a fingerprint of its 
contents, $f$, is computed using hashing techniques such as Rabin fingerprints~\cite{rabin81}. 
When this fingerprint meets a certain criteria,
% such as $f mod D = r$ where $D$, the divisor, 
%and $r$ are predefined values; 
the position of the corresponding window defines the boundary of the chunk. 
This process is repeated until the entire file has been completely broken down into chunks. Next, 
a cryptographic hash or chunk ID of the chunk is computed using a technique such as MD5 or SHA.
After a file is chunked, the index containing the chunk IDs of backed up chunks 
is queried to determine duplicate chunks. New chunks are written to disk and the 
index is updated with their chunk IDs. A file recipe containing all the information 
required during reconstruction is generated. The index also contains some metadata 
about each chunk, such as its size and disk location.
%The compression rate obtained by deduplication depends on the inherent content overlapping degree in a dataset, 
%the average size of chunks, and the chunking method\cite{poli04}. 
%In general, smaller chunks yield better deduplication.

Data deduplication has been applied as  a post-processing operation for offline removal
of redundant content in the file system data management once backup data is written in a temporary stage area.
Dynamic inline deduplication is an effort to improve storage efficiency  
by detecting and removing  duplicates before data is written to the backup storage,
thus reduces extra disk space required to hold temporary data. 
This approach requires a fingerprint  comparison of the new content chunks with
previously stored fingerprints. 
}
As data grows to be big, content chunk lookup for fingerprint comparison from disk storage
becomes too slow to be scalable.
%and searching such disk
%However, unless some form of locality or similarity is exploited, inline, chunk-based deduplication, 
%when done at a large scale faces what has been termed the disk bottleneck problem: to facilitate fast chunk ID lookup, 
%a single index containing the chunk IDs of all the backed up chunks must be maintained. 
%As the backed up data grows, the index overflows the amount of RAM available and must be paged to disk. 
%Without locality, the index cannot be cached effectively, and it is common for nearly 
%every index access to require a random disk access. This disk bottleneck severely limits deduplication throughput.
Several techniques have been proposed to speedup searching of duplicate 
content~\cite{bottleneck08,extreme_binning09,sparseindex09}. For example,  
Zhu et al.~\cite{bottleneck08} tackle it 
by using an in-memory Bloom filter and prefetch groups of chunk IDs that are likely to be 
accessed together with high probability. It takes significant memory resource for filtering.
NG et al.~\cite{ NGmiddleware2011}  use  
a related filtering technique for integrating deduplication in Linux  file system and the memory
consumed is up to 2GB for a single machine. That is still too big in our context discussed below. 
%Lillibridge et al.~\cite{sparseindex09} break list of chunks 
%into large segments, the chunk IDs in each incoming segment are sampled and the segment is 
%deduplicated by comparing with the chunk IDs of only a few carefully selected backed up segments. 
%These are segments that share many chunk IDs with the incoming segment with high probability.
%Deepavali et al.~\cite{extreme_binning09}  use signature-based file similarity  and group similar files
%into the same physical location (bins) to deduplicate against each other.

@@

There are multiple choices in designing a backup architecture  for VM snapshots.
We discuss the following design options with a consideration on their strengths and weakness.
\begin{enumerate}
\item  {\em An external and dedicated backup storage system.} 
In this architecture setting, a separate backup storage system using
the standard backup and deduplication techniques can be deployed~\cite{bottleneck08,extreme_binning09,sparseindex09}. 
This system is attached to the cloud network and every machine can periodically transfer snapshot data to 
the attached backup system. 
A key weakness of this approach is communication bottleneck between a large number of machines
in a cloud to this centralized  service.
Another weakness is that the cost of allocating separate resource for dedicated backup  can be expensive.
Since most of backup data is not used eventually, CPU and memory resource in such a backup cluster may not be fully utilized.
\item {\em A decentralized and co-hosted backup system with full deduplication.}
In this option, the backup system runs on an existing set of cluster machines.
% and a distributed storage architecture for backup allows a possible exploitation of  data locality between
%the source of data and storage location of backup data. 
The disadvantage is that 
%the backup service would compete CPU, memory, and disk resources with the other cloud services.
even such a  backup service may only use  a fraction of the existing disk storage, 
fingerprint-based search does require a significant amount of memory for fast lookup of identical content chunks.
This competes memory resource  with the existing applications.
%Decentralized deduplication is studied in ~\cite{Clements2009} and the focus is on
%block-level copy-on-write and compare-by-value techniques.

@@

With these considerations in mind, we propose a decentralized backup architecture with multi-level and selective 
deduplication. This service is hosted   in the existing set of machines and resource usage is controlled
with a minimal impact to the existing applications.
The duplication process is first conducted among data within each VM
and then is conducted  across VMs to utilize a common dataset.  
Given the concern that searching duplicates across  VMs  is a global feature which can affect parallel performance
and complicate failure management,
we would like to use a small common dataset while still maintaining a cost-effective deduplication ratio.
For this purpose, we exploit the data characteristics of snapshots and only store most popular data.
%
%in the  our studies show that there are a large amount of data unmodified in VM snapshots
%and shared among many users (e.g. OS data). The sharing pattern follows a zip-like distribution.
%With this characteristic, we can store a small amount of popular data items which coverage a large amount of
%snapshot data blocks. 
%We discuss our design and data analysis in the next section.

### 2012-03-01 vm_snapshot_sim/2012concl.tex

\section{Conclusion}
\label{sect:final}
In this paper we propose and present a multi-level selective deduplication scheme for VM snapshots. 
with minimal cost. Inner-VM deduplication localizes backup data dependence and exposes more parallelism  
while we use  data characteristic in VM cloud backup to identify popular common data blocks
and cover  a large percentage of snapshot content blocks wit a small amount of memory index.
Our evaluation using Aliyun's implementation has shown that 
level 1 and level 2 deduplication can reduce the storage need by 80\%. 
Level 3 with small CDS memory usage can accomplish close to 44\% of what full fingerprint-based
deduplication can do and bring additional 6\% storage reduction compared to the original data size.
Noted that 6\% is still significant, which is about 24GB per each VM and for a 1000 node Aliyun cluster,
this is about 600 terabytes.

>>>>

\section{Conclusion}
\label{sect:final}
The main contribution of this paper is a multi-level selective deduplication scheme among VM snapshots. 
Inner-VM deduplication localizes backup data dependence and exposes more parallelism  
while global deduplication using a small common data set appeared in OS and data disks
effectively  covers a large amount of snapshot blocks.
Our evaluation using Aliyun's data has shown that 
level 1 and level 2 deduplication can reduce the storage need by 78\% while level 2 inner 
fingerprint comparsion contributes 4.5\%%. 
Level 3 can add additional 10.5\% reduction and accomplish 75\% of what full fingerprint-based
deduplication can do. Our scheme uses a very small amount of memory on each node, and leaves
room for additional optimization we are further studying.
%Noted that 6\% is still significant, which is about 24GB per each VM and for a 1000 node Aliyun cluster,
%this is about 600 terabytes.

### 2012-03-01 vm_snapshot_sim/2012eval.tex

We have used a storage dataset of  1323 VM users collected from 100 Aliyun's cloud nodes  to evaluate the effectiveness.
In this dataset, there are 10 snapshots per each VM user and the total amount of space investigated 
is 23.5 terabytes.
Each VM file is  divided into a set of content blocks using
TTTD chunking~\cite{??} with an average size of 4KB. Each segment is of size 2MB.  
Popularity is computed by using 90\% of dataset, which reflects our setting that the system recomputes
CDS every 1-2 days to catch up the popularity trend.

@@

\begin{figure}
  \centering
  \epsfig{file=images/overall_effect.eps, width=3in}
  \caption{Impacts of 3-level deduplication. The height of each bar is the data size after 
deduplication divided by the original data size and the unit is percentage. }

@@

\begin{figure}
\centering
\epsfig{file=images/ay41a_big.data_disk.cdf.eps, width=3in}
%\epsfig{file=images/cdf.disk.v4.eps, width=3in}
%\epsfig{file=images/datadisk.count_rank.eps, width=3in}
%\epsfig{file=images/35os.count_rank.eps, width=3in}

@@

We further study how CDS covers content blocks used in image files.
Figure \ref{fig:userdatacoverage} shows the cumulative coverage  ratio of  popular common user data blocks. 
The $X$ axis is the rank of common user data blocks.
Let $S_i$ and $F_I$ be the size and duplicate count 
of the $i$-th block ranked by its duplicate rank.  Then $Y$ axis is the coverage of
the common dataset covering data items from rank 1 to rank $i$. Namely  
\[
\frac{ \sum_{i=1}^{i} S_i * F_i} {\mbox{Total data size}}
\]

@@

%We rank unique data blocks by their duplication count, and choose the hottest blocks as CDS. 
Figure \ref{fig:datacdssize} shows the relationship between CDS cache size and relative space saving ratio
compared to the full deduplication.  The unit of CDS size is gigabytes.
We define \emph{space saving ratio} as the space saving of the CDS method divided by 
full deduplication saving ratio. 
With a small amount of CDS data can still  accomplish more than 50\% of what perfect deduplication can do.
%But this effect decreases when more data are added to CDS. 
%The lower bound of CDS space saving ratio is 50\%, which is very easy to accomplish.

\begin{figure}
  \centering
  \epsfig{file=images/uniquedata-saving.eps, width=3in}
  \caption{Relative ratio of space size compared to full deduplication when CDS size changes.}
  \label{fig:datacdssize}
\end{figure}

CDS size is restricted by system memory resource.
Figure \ref{fig:datacds} shows how CDS space saving is affected by the system scale in terms of
the dataset size. 
In this experiment we first set out a goal of space saving ratio, 
then watch how much data needs to be placed in CDS cache to achieve this goal.
From the graph we can see a 75\% saving goal lead to a stable ratio between 
CDS size and data size, which requires 0.01\% of data to be put in CDS.

Base on above data we can estimate the size of data CDS and its effect. 
Currently we prepared 500MB memory per machine to store CDS meta, then it can represent 50GB of data. 
If we assume each VM has 30GB of user data at runtime, and we host 25 VMs per machine, 
 maintain 10 snapshots per VM, each brings 10\% additional modified data. 
Thus the user data in snapshot system is 1.5TB per machine. So the upper bound of 
$CDS size/ Data size = 0.033$, which is sufficient for the 75\% saving goal.

\begin{figure}
  \centering
  \epsfig{file=images/cds_scale.0.7.eps, width=3in}
  \caption{The relative size ratio of  CDS over the data size. }
  \label{fig:datacds}
\end{figure}

>>>>

We have used a storage dataset of  1323 VM users collected from 100 Aliyun's cloud nodes  to evaluate the effectiveness.
In this dataset, there are 10 snapshots per each VM user and the total amount of space investigated 
is 23.5 terabytes.
Each VM file is  divided into content blocks of
variable sizes~\cite{similar94,rabin81} with an average size of 4KB. Each segment is of size 2MB.  
Popularity is computed by using 90\% of dataset, which reflects our setting that the system recomputes
CDS every 1-2 days to catch up the popularity trend.

@@

\begin{figure}
  \centering
  \epsfig{file=images/overall_effect.eps, width=2.8in}
  \caption{Impacts of 3-level deduplication. The height of each bar is the data size after 
deduplication divided by the original data size and the unit is percentage. }

@@

\begin{figure}
\centering
\epsfig{file=images/ay41a_big.data_disk.cdf.eps, width=2.8in}
%\epsfig{file=images/cdf.disk.v4.eps, width=3in}
%\epsfig{file=images/datadisk.count_rank.eps, width=3in}
%\epsfig{file=images/35os.count_rank.eps, width=3in}

@@

We further study how CDS covers content blocks used in image files.
Figure \ref{fig:userdatacoverage} shows the cumulative coverage  ratio of  popular common user data blocks. 
The $X$ axis is the rank of common user data blocks.
Let $S_i$ and $F_I$ be the size and duplicate count 
of the $i$-th block ranked by its duplicate rank.  Then $Y$ axis is the coverage of
the common dataset covering data items from rank 1 to rank $i$. Namely  
\[
\frac{ \sum_{i=1}^{i} S_i * F_i} {\mbox{Total data size}}.
\]

Thus with about 1\% of blocks on data disks, CDS can cover about 38\% of total data blocks
appeared in in all data snapshots. CDS index uses about 150MB memory per machine.

@@

%We rank unique data blocks by their duplication count, and choose the hottest blocks as CDS. 
Figure \ref{fig:datacdssize} shows the relationship between CDS cache size and relative space saving ratio
compared to the full deduplication.  The unit of CDS size is gigabytes.
We define \emph{space saving ratio} as the space saving of the CDS method divided by 
full deduplication saving ratio. 
With a 100GB CDS data (namely 1GB CDS index) can still  accomplish about 75\% 
of what perfect deduplication can do.
%But this effect decreases when more data are added to CDS. 
%The lower bound of CDS space saving ratio is 50\%, which is very easy to accomplish.

\begin{figure}
  \centering
  \epsfig{file=images/uniquedata-saving.eps, width=2.8in}
  \caption{Relative ratio of space size compared to full deduplication when CDS size changes.}
  \label{fig:datacdssize}
\end{figure}

CDS size is restricted by system memory resource.
Figure \ref{fig:datacds} shows how CDS space saving ratio
compared to the full deduplication is affected by the dataset size. 
The calculation is done after level 1 and level 2 deduplication.
$X$ axis uses the data size collected  from up to 1190 VMs, which occupies about
47 machines. 
In this experiment we first set out a goal of space saving ratio completed, 
then watch how much data needs to be placed in CDS cache to achieve this goal.
From the graph we can see a 75\% saving ratio lead to a stable ratio between 
CDS size and data size, which requires 1\% of data to be placed in CDS.

When we deal with a cluster of 1,000 nodes, we expect that using
1\% of data disks as the CDS size can cover more what we have seen from this
1190 VM dataset, assuming that the average behavior of every subcluster with 47 machines
exhibits a similar commonality. 
In addition to this, CDS for OS disks will become even more effective when
there are more VMs sharing the same collection of OS releases. 
%Base on above data we can estimate the size of data CDS and its effect. 
%Currently we prepared 500MB memory per machine to store CDS meta, then it can represent 50GB of data. 
%If we assume each VM has 30GB of user data at runtime, and we host 25 VMs per machine, 
% maintain 10 snapshots per VM, each brings 10\% additional modified data. 
%Thus the user data in snapshot system is 1.5TB per machine. So the upper bound of 
%$CDS size/ Data size = 0.033$, which is sufficient for the 75\% saving goal.

\begin{figure}
  \centering
  \epsfig{file=images/cds_scale.0.7.eps, width=2.8in}
  \caption{The relative size ratio of  CDS over the data size. }
  \label{fig:datacds}
\end{figure}
